Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms

### 2013-07-23 CGP Journal/main.tex
Added my image to my biography section
Final content edit, now its time to shrink things

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
Understanding how search operators interact with solution representation is
a critical step to improving search.
In Cartesian Genetic Programming (CGP), and Genetic Programming (GP)
in general, the complex genotype to phenotype map makes achieving this understanding a
challenge.  By examining tuned parameter values, the search quality of CGP variants at different
problem difficulties, node behavior, and offspring replacement properties we
seek to comprehend observed characteristics of CGP search.  Our focus is twofold:
methods to prevent wasted CGP evaluations (\emph{Skip}, \emph{Accumulate},
and \emph{Single}) and methods to overcome CGP's search limitations imposed by
genome ordering (\emph{Reorder} and \emph{DAG}).

Our results show that most CGP techniques evolve genomes that
are highly inactive, very redundant, and full of seemingly useless constants.
On some tested problems we found less than 1\% of the genome was actually required to
encode the evolved solution.
Without \emph{Reorder} or \emph{DAG} almost the entire latter portion of the genome
is never used by any ancestor of the evolved solution.  More generally, our
results suggest that \emph{Skip-Reorder} and \emph{Single-Reorder}
are most likely to solve hard problems in the least number of evaluations and
the least amount of time with the least amount of degenerate
behavior.

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{W}{hile effective} evolutionary methods are often simple to design
and implement, discerning the reason that system
leads to effective search can be challenging.
This is exceptionally
true for Genetic Programming (GP), which often employs nontrivial
genotype to phenotype maps and operators that act on genotypes without regard to
phenotypic impact.  Yet understanding the root causes of evolutionary success
and failure is critical to improving existing techniques, designing new techniques,
and understanding how and where to apply each optimization system.

@@

In this work we set out to develop a deeper understanding of what makes Cartesian
Genetic Programming (CGP) an effective evolutionary optimizer.
A few studies of CGP's evolutionary mechanisms have
been done previously, related to bloat~\cite{miller:2001:bloat},
neutrality~\cite{vassilev:2000:neutrality}, and structural bias~\cite{payne:2009:bias}.
We shall focus on the interaction between mutation and genome ordering.

@@

CGP represents DAGs using a linear genome of integer values.  Each node in the
DAG is encoded as a tuple of genes, with one gene specifying the function
the node applies to its inputs, and the remaining genes expressing where the
node takes input from.  Nodes can take input from either a problem input or
any node preceding them in the linear genome.  Restricting connections in this way
prevents the creation of cycles, while still allowing CGP to reuse values.  This is
in contrast to tree based GP, which must recalculate a value everywhere it is needed.
To complete the representation, a set
of extra genes are included at the end of the genome to specify which nodes or
input locations to use as function outputs.

As both output locations and information flow in the DAG are evolvable, often
only a tiny fraction of the genome participates in creating the
output values.  These nodes are referred to as ``active,'' with the nodes not
being used to create output values referred to as ``inactive.''  Inactive nodes
allow CGP individuals to drift genetically, as they can be mutated without effecting
the fitness of the individual.  This genetic drift can then be incorporated, as
mutation can change the DAG structure causing previously inactive nodes to become
active.  Previous work suggests CGP is most efficient when up to 95\% of the genome
is inactive~\cite{miller:2006:redundancy}, while more recent work suggests this may
be a result of hidden parsimony pressure in CGP~\cite{goldman:2013:ordering}.

CGP uses very simple evolutionary mechanisms.  The most common evolutionary strategy
is $\mu + \lambda$ where $\mu \leftarrow 1$ and $\lambda \leftarrow 4$.
This means each generation a single parent produces four offspring using
mutation.  The best offspring then competes with the parent,\footnote{Ties are broken
randomly.} with the offspring
replacing the parent if it is no less fit.  This replacement strategy encourages
neutral drift.  Mutation
is commonly performed by mutating each gene of each node at a set probability, where mutation
involves changing a gene randomly to some different valid value.  For example,
if a function gene is chosen for mutation, its new value is randomly chosen from
all possible functions, with the exception of the genes current value.  Combined,
this form of CGP can be viewed as a stochastic hill climber with neutrality.
For a more in depth description of CGP, see~\cite{miller:2011:chapter2}.

\section{Duplicate Evaluation Avoidance}
\label{sec:duplicate}
CGP's mutation operator is traditionally applied to both active and inactive genes uniformly.
This has the
potential to create offspring who's only mutated genes are in inactive nodes.
These offspring are actively identical (contain identical active genes) to their
parents, and therefore by definition have identical fitness to their parents.
As such, re-evaluating these individuals is computationally wasteful, as their fitness
is known.
Detecting duplicated individuals can be done without significant overhead as
finding the set of active nodes is already done prior to
evaluation~\cite{vasicek:2012:efficient}.
Previous work~\cite{goldman:2013:cgpwaste} has taken an initial look at the
effect this waste can have, and proposed \emph{Skip}, \emph{Accumulate}, and
\emph{Single} as methods for avoiding or preventing these duplicate evaluations.

@@

Because \emph{Skip} does not modify any evolutionary mechanisms found in traditional CGP,
it will produce identical individuals, with a potential reduction in evaluations.
If the mutation rate is high enough relative
to the number of active genes, \emph{Skip} will require the same number of evaluations as traditional CGP. This is because
the probability of an offspring being actively identical to its parent is effectively zero.
When the mutation rate is low enough relative to the number of active genes, some number of
offspring will be actively identical to their parents, resulting in a reduction in evaluations
but no change in the evolutionary trajectory.

We would expect both intuitively and from initial
experimentation~\cite{goldman:2013:cgpwaste} that \emph{Skip} will be less sensitive
to the mutation rate than normal CGP.  We would also expect \emph{Skip} to use a lower
mutation rate than normal CGP when both are optimized to reduce evaluations.
The former comes from the fact that in normal CGP any mutation rate that has a significant
probability of creating offspring actively identical to their parents wastes
evaluations.  In \emph{Skip} there are only two penalties for reducing the
mutation rate.  From an evolutionary standpoint, if search becomes trapped in a local optima, low mutation
rates may have increased difficulty escaping to find the global optimum.  From a performance standpoint,
exceptionally low mutation rates may spend a prohibitive amount of time
attempting to produce an evaluable offspring.
Note that these penalties also exist in normal CGP.

@@

This modification gives \emph{Single} three properties distinct from the other
forms of duplicate avoidance.  First, \emph{Single} forces offspring to have
an active gene which differs from their parent.  This limits drift as the
mutated active gene must either be to an intron or represent another way to
code a solution of equal quality.  Second, as a benefit of forced changes
\emph{Single} avoids the overhead of repeatedly generating individuals without
creating an evaluable offspring.  Third, \emph{Single} does not require a mutation
rate parameter, effectively setting the mutation rate to $\frac{1}{a}$ for active
genes and $\frac{1}{a+1}$ for inactive genes, where $a$ is the number of active genes.

@@

Ordering, when combined with the mutation's uniform resetting of connection genes, may be the cause of CGP's
immensely inactive genomes.  As was rigorously defined in~\cite{goldman:2013:ordering},
the number of active genes in a genome is expected to scale logarithmically
with genome size, independent of problem application.  Furthermore, the less
active a genome, the lower the likelihood of structure ordering preventing structure
connection.

As an extension to the theory presented in~\cite{goldman:2013:ordering}, we have
derived the formula which predicts the probability a node is active based on its
index, given in Equation~\ref{eq:prob} and shown in Figure~\ref{fig:prob}.
\begin{equation}
p(i) \leftarrow 1 - (\frac{N+I-1}{N+I})^O \cdot \prod_{j=i+1}^{N} 1 - p(j) \cdot (1-(\frac{j+I-1}{j+I})^A)
\label{eq:prob}
\end{equation}
In this equation, $i$ is the index of the working node, $N$ is the genome size, $I$ is the number of input locations,
$O$ is the number of output locations, and $A$ is the arity of each node.
This equation is constructed as a negation, such that we determine the probability
the node is not active due to each potential source that could make it active.
The product term calculates the probability that the node at index $j$ is connected to the
node at index $i$, multiplied by the probability that the node at index $j$ is active.  This is
negated to determine the probability node $i$ is not active because of node $j$.
All together, the product term calculates the probability that no active node is
connected to the node at index $i$.  Nodes can also become active if they are directly
connected to by an output location.  As such the product is scaled by the probability
that all output locations connect to inputs or nodes not at index $i$.  This
formula assumes only operator bias, not selective pressure, is acting on the
genome.

@@

\emph{Reorder} is used once each generation to shuffle the nodes of the parent's
genome.  As the shuffling does not semantically change the parent, it does
not require re-evaluated.  Furthermore, as shuffling does not change the requirement
that nodes only depend on those preceding them in the genome, shuffling does not
require any changes to any other CGP methods, such as evaluation and mutation.
It is important to note, however, that the potential mutations that can be applied to the parent
have changed.  Consider two nodes $X$ and $Y$ which are in the genome, and
have no dependence relationship.  As $X$ does not take input from $Y$ and
$Y$ does not take input from $X$, either can be placed before the other
in the genome.  Without shuffling, the initial ordering of nodes is fixed.
No operation in \emph{Normal} is capable of making $X$
take input from $Y$.  \emph{Reorder}'s shuffling can allow
subsequent mutations to make the connection.

@@

The price \emph{Reorder} pays for this potential improvement is increased mutational search
space.  Though a shuffled individual is phenotypically unchanged,
its mutational adjacency is significantly different.  As such the variety of offspring
an individual can create is increased when using \emph{Reorder}.  While this increase
makes finding a specific offspring that improves upon the parent less likely, the
general likelihood of producing an offspring that improves upon the parent may 
more than offset this drawback.

\subsection{DAG}
\label{sec:dag}
\emph{Normal} and \emph{Reorder} use node location in a linear genome to create
an efficient protection against cycles: nodes must connect forward.  \emph{DAG}
removes this scheme of location, relying on more complex methodologies to
prevent cycles.  To accomplish this goal, modifications must be made to how CGP
performs mutation and evaluation.

As in \emph{Normal} and \emph{Reorder}, when a connection gene is chosen for mutation, its value is chosen randomly.
However, the options to choose from are only limited to those connections which do not
create cycles.
To achieve this, we incrementally determine
which nodes transitively depend on the mutating node and which do not, with
information stored in the $known$ map.  This data structure is initialized with
the information that the node is dependent on itself and that all input locations
are not dependent on the mutating node.  Nodes are then
tested in a random order for dependence using the \Call{IsDependent}{} algorithm given in Figure~\ref{fig:dag}.
The first node returned at the top level that is not dependent on the mutating node is then
used as the new connection gene value.  As a result, mutating a connection gene
cannot introduce a cycle.

@@

The method for evaluating \emph{DAG} individuals is very similar to the method for
efficiently evaluating \emph{Normal} and \emph{Reorder} individuals.  As before,
preprocessing is done to determine which nodes are active, and only those nodes are
executed during evaluation.  Also as before, the process begins from the output locations,
recursively following connection genes and marking nodes as active until the input locations
are reached.  At this point, the process changes.  In \emph{DAG}, we must determine not only the set of
nodes that are active, but the order in which those nodes should be executed.

@@

While \emph{Reorder} and \emph{DAG} both reduce mutational limitations, \emph{DAG}'s changes are more sweeping.
Though both can create any connect that does not form a cycle,
\emph{Reorder} maintains a bias toward certain kinds of connections.
Consider again two nodes $X$ and $Y$
such that neither is dependent on the other.  If $X$ is transitively dependent on very
few nodes, and $Y$ is transitively dependent on many nodes, \emph{Reorder} is more likely
to shuffle $X$ to precede $Y$.  This is true because at any point during the construction
of the new ordering, the probability
that all of $X$'s dependencies have been added is
higher than for $Y$'s dependencies.  As a result we expect $X$ to be put into $addable$ sooner,
and therefore we expect it to be added sooner.  The greater the discrepancy in the dependency
set sizes the less likely the order between the nodes will be reversed.
Conversely, \emph{DAG} has no such bias.  It is always possible to mutate $X$ to connect to $Y$.

The probability of connecting $X$ to $Y$ is also influenced by how many nodes transitively depend $X$ and $Y$.
For example, assume $X$ has more nodes transitively dependent on it than $Y$.
In \emph{Reorder}, $X$ is again more likely to be placed before $Y$, this time because the number
of possible nodes preceding it is reduced.  \emph{DAG} is biased in the reverse, such that $X$ will
have a higher chance of connecting to $Y$ that $Y$ to $X$.  This is because $X$ can connect to fewer
nodes, so each mutation has a higher chance of choosing $Y$.

@@

All told we have proposed three new methods for avoiding duplicate evaluations
(\emph{Skip}, \emph{Accumulate}, and \emph{Single}) and two new methods for
genome ordering (\emph{Reorder} and \emph{DAG}).
See Table~\ref{tab:glossary} for a short description each variant.
As these two
sets are non overlapping, and there is no intuitive groupings, we chose to test
all possible combinations in our empirical analysis.  As a control we included \emph{Normal} genome ordering,
as it provides insight to how each of the duplicate prevention techniques work
if applied alone.  Similar measures are not needed in testing the genome ordering techniques,
as \emph{Skip} has no impact on evolutionary search, and is a strict improvement
to using no duplication detection at all.

In total this creates 9 algorithm combinations.  Of these, previous work
has been reported showing the effectiveness of the duplication methods with \emph{Normal} ordering~\cite{goldman:2013:cgpwaste}
as well as analysis of all three ordering methods with \emph{Single}~\cite{goldman:2013:ordering}.
This means there remain 4 novel combinations to be tested.
This is also the first time the ordering methods
have been rigorously tested for quality, as much of the work in~\cite{goldman:2013:ordering}
was analysis oriented.

@@

\subsection{Parameter Setting}
\label{sec:parameter}
To perform a fair comparison, we ensure proper parameter configuration
to avoid potential bias that might benefit
a specific method.  While tuning can lead to the alternative issue
of methods only being effective after extensive problem specific tuning, we set out to use
rough grained values and provide each technique with equal tuning time to alleviate that bias.

@@

	\end{tabular}
	\caption{Results for algorithm configuration on each problem.  Highlighted row
	is the control configuration \emph{Skip-Normal}.  MES=Median Evaluations to Success; MAD=Median
	Absolute Deviation; p-value=Result from Mann-Whitney~U comparison with control.
	}
	\label{tab:results}
\end{table*}

@@

Of the 12 configurations including \emph{Single}, 11 used a genome size smaller
than all configurations not using \emph{Single}.  The exception was \emph{Single-Normal}
on Multiply.  No other algorithm had as much impact on genome size tuning as
the inclusion or exclusion of \emph{Single}.  In previous work~\cite{goldman:2013:ordering}
it appeared that \emph{Reorder} and \emph{DAG} allowed CGP to use smaller genomes.
In light of our results, this conclusion should be amended.
It appears that \emph{Single} is most effective when the genome size is small and
that \emph{Reorder} and \emph{DAG} are best able to cope with smaller genome sizes.
Independent of duplication method, the ordering methods do not seem to create any
pressure on which genome size is chosen.

Unlike genome size, all configurations choose approximately the same mutation rate,
with different rates for each problem.  With the exception of Parity, there appears
to be a general consensus that each gene should be mutated with probability 0.002,
independent of genome size.  Parity used a higher mutation rate, but this is likely
because 3-bit Parity is by far the easiest problem tested and therefore likely benefits
from higher exploration.

On many of the problems, many of the configurations have little to no statistical
difference from the control in their evaluations required to reach the global optimum.
On Parity and Encode it is probably safe to conclude that \emph{Single} is worse than
the control.  As these are the easiest problems tested that difference may have
little real world meaning.  \emph{DAG} is almost certainly worse than the control on
Decode.  Its results are so much worse that we can also likely conclude that \emph{DAG}
is worse than all non \emph{DAG} configuration on that problem, even though no statistical
tests are presented here.  The control on multiply did by far the worst of all configurations,
with only \emph{Accumulate Normal} likely failing statistical significance.

By ordering the algorithm combinations by MES and examining that ordering across each
problem, we can look for general trends.  From the table, it appears that on easier problems
the method of duplication used is the strongest predictor of rank, while
the ordering methods are more predictive on harder problems.  This comes from the
observation that in for Parity and Encode (Upper half of Table~\ref{tab:results})
the first column is effectively sorted, while for Decode and Multiply
(Lower half of Table~\ref{tab:results})
the second column is sorted.
Each table includes coincidentally one exception to these rules.  On Parity and
Encode, pairings with \emph{DAG} are ranked lower than other ordering methods, so much
so as to break the ordering once on both problems.  On Decode \emph{Accumulate-Reorder}
ranks worse than expected.  Finally on Multiply \emph{Single-DAG} does far better
than expected.

@@

So now the question is why does \emph{Normal} use so little of its genome?  To
answer this we must consider what can prevent a node from becoming active.  For
a node to never be active, all tested offspring with mutations that result in that node being activated
must have been less fit than their parent.  Even if those offspring are no more fit, the selection method
used by CGP would have allowed that offspring to replace its parent, at least probabilistically.  This leads
to two possible causes for a node never being active: bad genes or infrequently tried.

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
Understanding how search operators interact with solution representation is
a critical step to improving search.
In Cartesian Genetic Programming (CGP), and Genetic Programming (GP)
in general, the complex genotype to phenotype map makes achieving this understanding a
challenge.  By examining aspects such as tuned parameter values, the search quality of CGP variants at different
problem difficulties, node behavior, and offspring replacement properties we
seek to better understand the characteristics of CGP search.  Our focus is twofold:
creating methods to prevent wasted CGP evaluations (\emph{Skip}, \emph{Accumulate},
and \emph{Single}) and creating methods to overcome CGP's search limitations imposed by
genome ordering (\emph{Reorder} and \emph{DAG}).

Our results show that most CGP techniques evolve genomes that
are highly inactive, very redundant, and full of seemingly useless constants.
On some tested problems we found less than 1\% of the genome was actually required to
encode the evolved solution.
Furthermore, traditional CGP ordering results in large portions of the genome
that are never used by any ancestor of the evolved solution.
\emph{Reorder} and \emph{DAG} allow evolution to utilize the entire genome.
More generally, our
results suggest that \emph{Skip-Reorder} and \emph{Single-Reorder}
are most likely to solve hard problems using the least number of evaluations and
the least amount of time while better avoiding degenerate
behavior.

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{W}{hile effective} evolutionary methods are often simple to design
and implement, discerning the reason that system
leads to effective search can be challenging.
This is exceptionally
true for Genetic Programming (GP), which often employs both nontrivial
genotype to phenotype maps and operators that act on genotypes without regard to
phenotypic impact.  Yet understanding the root causes of evolutionary success
and failure is critical to improving existing techniques, designing new techniques,
and understanding how and where to apply each optimization system.

@@

In this work we set out to develop a deeper understanding of what makes Cartesian
Genetic Programming (CGP) an effective evolutionary optimizer.
Previous studies of CGP's evolutionary mechanisms have
focused on bloat~\cite{miller:2001:bloat},
neutrality~\cite{vassilev:2000:neutrality}, and structural bias~\cite{payne:2009:bias}.
We shall focus on the interaction between mutation and genome ordering.

@@

CGP represents DAGs using a linear genome of integer values.  Each node in the
DAG is encoded as a tuple of genes, with one gene specifying the function
the node applies to its inputs, and the remaining genes expressing where the
node takes input from.  Nodes can take input from either a problem input or
any node preceding them in the linear genome.  Restricting connections in this way
prevents the creation of cycles, while still allowing CGP to reuse values.  This is
in contrast to tree based GP, which must duplicate functionality everywhere the same value is needed.
To complete the representation, a set
of extra genes are included at the end of the genome to specify which nodes or
input locations to use as function outputs.

As both output locations and information flow in the DAG are evolvable, often
only a tiny fraction of the genome participates in creating the
output values.  These nodes are referred to as ``active,'' with the nodes not
being used to create output values referred to as ``inactive.''  Inactive nodes
allow CGP individuals to drift genetically, as they can be mutated without affecting
the fitness of the individual.  This genetic drift can then be incorporated, as
mutation can change the DAG structure causing previously inactive nodes to become
active.  Previous work suggests CGP is most efficient when up to 95\% of the genome
is inactive~\cite{miller:2006:redundancy}, while more recent work suggests this may
be a result of hidden parsimony pressure in CGP~\cite{goldman:2013:ordering}.

CGP uses very simple evolutionary mechanisms.  The most common evolutionary strategy
is $\mu + \lambda$ where $\mu \leftarrow 1$ and $\lambda \leftarrow 4$.
This means during each generation a single parent produces four offspring using
mutation.  The best offspring then competes with the parent,\footnote{Ties are broken
randomly.} with the offspring
replacing the parent if it is no less fit.  This replacement strategy encourages
neutral drift.  Mutation
is commonly performed by mutating each gene of each node at a set probability, where mutation
involves changing a gene randomly to some different valid value.  For example,
if a function gene is chosen for mutation, its new value is randomly chosen from
all possible functions, excluding the gene's current value.  Combined,
this form of CGP can be viewed as a stochastic hill climber with neutrality.
For a more in depth description of CGP, see~\cite{miller:2011:chapter2}.

\section{Duplicate Evaluation Avoidance}
\label{sec:duplicate}
CGP's mutation operator is traditionally applied to both active and inactive genes uniformly.
This has the
potential to create offspring who's only mutated genes are in inactive nodes.
These offspring are actively identical (contain identical active genes) to their
parents, and therefore by definition have identical fitness to their parents.
As such, re-evaluating these individuals is computationally wasteful as their fitness
is known.
Detecting duplicated individuals can be done without significant overhead since
finding the set of active nodes is already done prior to
evaluation~\cite{vasicek:2012:efficient}.
Our previous work~\cite{goldman:2013:cgpwaste} has taken an initial look at the
effect this waste can have, and proposed \emph{Skip}, \emph{Accumulate}, and
\emph{Single} as methods for avoiding or preventing these duplicate evaluations.

@@

Because \emph{Skip} does not modify any evolutionary mechanisms found in traditional CGP,
it will produce identical individuals, with a potential reduction in evaluations.
If the mutation rate is high enough relative
to the number of active genes, \emph{Skip} will require the same number of evaluations as traditional CGP, since
the probability of an offspring being actively identical to its parent is effectively zero.
When the mutation rate is low enough relative to the number of active genes, some number of
offspring will be actively identical to their parents, resulting in a reduction in evaluations
but no change in the evolutionary trajectory.

We would expect both intuitively and from our initial
experimentation~\cite{goldman:2013:cgpwaste} that \emph{Skip} will be less sensitive
to the mutation rate than normal CGP.  We would also expect \emph{Skip} to use a lower
mutation rate than normal CGP when both are optimized to reduce evaluations.
The former comes from the fact that in normal CGP any mutation rate that has a significant
probability of creating offspring actively identical to their parents wastes
evaluations.  In \emph{Skip} there are only two penalties for reducing the
mutation rate.  From an evolutionary standpoint, if search becomes trapped in a local optima, low mutation
rates may have increased difficulty escaping to find the global optimum.  From a performance standpoint,
exceptionally low mutation rates may spend a prohibitive amount of time
attempting to produce an evaluable offspring.
Note that these penalties also exist in normal CGP.

@@

This modification gives \emph{Single} three properties distinct from the other
forms of duplicate avoidance.  First, \emph{Single} forces offspring to have
an active gene which differs from their parent.  This limits drift as the
mutated active gene must either be to an intron or represent another way to
code a solution of equal quality.  Second, as a benefit of forced changes,
\emph{Single} avoids the overhead of repeatedly generating individuals without
creating an evaluable offspring.  Third, \emph{Single} does not require a mutation
rate parameter, effectively setting the mutation rate to $\frac{1}{a}$ for active
genes and $\frac{1}{a+1}$ for inactive genes, where $a$ is the number of active genes.

@@

Ordering, when combined with the mutation's uniform resetting of connection genes, may be the cause of CGP's
immensely inactive genomes.  As was rigorously defined in our work~\cite{goldman:2013:ordering},
the number of active genes in a genome is expected to scale logarithmically
with genome size, independent of problem application.  Furthermore, the less
active a genome, the lower the likelihood of structure ordering preventing structure
connection.

As an extension to the theory presented in~\cite{goldman:2013:ordering}, we have
derived the formula which predicts the probability a node is active based on its
index, given in Equation~\ref{eq:prob} and plotted in Figure~\ref{fig:prob}.
\begin{equation}
p(i) \leftarrow 1 - (\frac{N+I-1}{N+I})^O \cdot \prod_{j=i+1}^{N} 1 - p(j) \cdot (1-(\frac{j+I-1}{j+I})^A)
\label{eq:prob}
\end{equation}
In this equation, $i$ is the index of the working node, $N$ is the genome size, $I$ is the number of input locations,
$O$ is the number of output locations, and $A$ is the arity of each node.
This equation is constructed as a negation, such that we determine the probability
the node is not active due to each potential source that could make it active.
The product term calculates the probability that the node at index $j$ is connected to the
node at index $i$, multiplied by the probability that the node at index $j$ is active.  This is
negated to determine the probability node $i$ is not active because of node $j$.
All together, the product term calculates the probability that no active node is
connected to the node at index $i$.  Nodes can also become active if they are directly
connected to by an output location.  As such the product is scaled by the probability
that all output locations connect to inputs or nodes not at index $i$.  This
formula assumes only operator bias, not selective pressure, is acting on the
genome.

@@

\emph{Reorder} is used once each generation to shuffle the nodes of the parent's
genome.  As the shuffling does not semantically change the parent, it does
not require re-evaluation.  Furthermore, as shuffling does not change the requirement
that nodes only depend on those preceding them in the genome, shuffling does not
require any changes to any other CGP methods, such as evaluation and mutation.
It is important to note, however, that the potential mutations that can be applied to the parent
have changed.  Consider two nodes $X$ and $Y$ which are in the genome, and
have no dependent relationship.  As $X$ does not take input from $Y$ and
$Y$ does not take input from $X$, either can be placed before the other
in the genome.  Without shuffling, the initial ordering of nodes is fixed.
No operation in \emph{Normal} is capable of making $X$
take input from $Y$.  \emph{Reorder}'s shuffling can allow
subsequent mutations to make the connection.

@@

The price \emph{Reorder} pays for this potential improvement is a larger mutational search
space.  Though a shuffled individual is phenotypically unchanged,
its mutational adjacency can be significantly different.  As such the variety of offspring
an individual can create is increased when using \emph{Reorder}.  While this increase
makes finding a specific offspring that improves upon the parent less likely, the
general likelihood of producing an offspring that improves upon the parent may 
more than offset this drawback.

\subsection{DAG}
\label{sec:dag}
\emph{Normal} and \emph{Reorder} use node location in a linear genome to create
an efficient protection against cycles: nodes must connect forward.  \emph{DAG}
removes this restriction and allows forward and backward connections, relying on other, more complex, methodologies to
prevent cycles.  To accomplish this goal, modifications must be made to how CGP
performs mutation and evaluation.

As in \emph{Normal} and \emph{Reorder}, when a connection gene is chosen for mutation, its value is chosen randomly.
However, the options to choose from are only limited to those connections which do not
create cycles.
To achieve this, we incrementally determine
which nodes transitively depend on the mutating node and which do not, with
information stored in a $known$ map data structure.  This data structure is initialized with
the information that the node is dependent on itself and that all input locations
are not dependent on the mutating node.  Nodes are then
tested in a random order for dependence using the \Call{IsDependent}{} algorithm given in Figure~\ref{fig:dag}.
The first node returned at the top level that is not dependent on the mutating node is then
used as the new connection gene value.  As a result, mutating a connection gene
cannot introduce a cycle.

@@

The method for evaluating \emph{DAG} individuals is very similar to the method for
efficiently evaluating \emph{Normal} and \emph{Reorder} individuals.  As before,
preprocessing is done to determine which nodes are active, and only those nodes are
executed during evaluation.  Also as before, the process begins from the output locations,
recursively following connection genes and marking nodes as active until the input locations
are reached.  At this point, the process changes.  In \emph{DAG}, we must determine not only the set of
nodes that are active, but also the order in which those nodes should be executed.

@@

While \emph{Reorder} and \emph{DAG} both reduce mutational limitations, \emph{DAG}'s changes are more sweeping.
Though both can create any connection that does not form a cycle,
\emph{Reorder} maintains a bias toward certain kinds of connections.
Consider again two nodes $X$ and $Y$
such that neither is dependent on the other.  If $X$ is transitively dependent on very
few nodes, and $Y$ is transitively dependent on many nodes, \emph{Reorder} is more likely
to shuffle $X$ to precede $Y$.  This is true because at any point during the construction
of the new ordering, the probability
that all of $X$'s dependencies have been added is
higher than for $Y$'s dependencies.  As a result we expect $X$ to be put into $addable$ sooner,
and therefore we expect it to be added sooner.  The greater the discrepancy in the dependency
set sizes the less likely the order between the nodes will be reversed.
Conversely, \emph{DAG} has no such bias.  It is always possible to mutate $X$ to connect to $Y$.

The probability of connecting $X$ to $Y$ is also influenced by how many nodes transitively depend on $X$ and $Y$.
For example, assume $X$ has more nodes transitively dependent on it than $Y$.
In \emph{Reorder}, $X$ is again more likely to be placed before $Y$, this time because the number
of possible nodes preceding it is reduced.  \emph{DAG} is biased in the reverse, such that $X$ will
have a higher chance of connecting to $Y$ that $Y$ to $X$.  This is because $X$ can connect to fewer
nodes, so each mutation has a higher chance of choosing $Y$.

@@

All told we have proposed three new methods for avoiding duplicate evaluations
(\emph{Skip}, \emph{Accumulate}, and \emph{Single}) and two new methods for
genome ordering (\emph{Reorder} and \emph{DAG}).
See Table~\ref{tab:glossary} for a short description each variant.
As these two
sets are non overlapping, and there is no intuitive groupings, we chose to test
all possible combinations in our empirical analysis.  As a control we included \emph{Normal} genome ordering,
as it provides insight to how each of the duplicate prevention techniques work
if applied alone.  Similar measures are not needed in testing the genome ordering techniques,
as \emph{Skip} has no impact on evolutionary search, and is strictly an improvement
to using no duplication detection at all.

In total this creates 9 algorithm combinations.  Of these, our previous work
has been reported showing the effectiveness of the duplication methods with \emph{Normal} ordering~\cite{goldman:2013:cgpwaste}
as well as analysis of all three ordering methods with \emph{Single}~\cite{goldman:2013:ordering}.
This means there remain 4 novel combinations to be tested.
This is also the first time the ordering methods
have been rigorously tested for quality, as much of the work in~\cite{goldman:2013:ordering}
was analysis oriented.

@@

\subsection{Parameter Setting}
\label{sec:parameter}
To perform a fair comparison, we ensure proper parameter configuration
to avoid potential bias that might benefit
a specific method.  While tuning can lead to the alternative issue
of methods only being effective after extensive problem specific tuning, we set out to use
rough-grained values and provide each technique with equal tuning time to alleviate that bias.

@@

	\end{tabular}
	\caption{Results for algorithm configuration on each problem.  Highlighted row
	is the control configuration \emph{Skip-Normal}.  MES=Median Evaluations to Success; MAD=Median
	Absolute Deviation; Active=Median number of active nodes in evolved solution;
	Used=Median number of nodes active in evolved solution after applying SIMPLIFY (See Section~\ref{sec:behavior});
	p-value=Result from Mann-Whitney~U comparison with control.
	}
	\label{tab:results}
\end{table*}

@@

Of the 12 configurations including \emph{Single}, 11 used a genome size smaller
than all configurations not using \emph{Single}.  The exception was \emph{Single-Normal}
on Multiply.  No other algorithm had as much impact on genome size tuning as
the inclusion or exclusion of \emph{Single}.  In our previous work~\cite{goldman:2013:ordering}
it appeared that \emph{Reorder} and \emph{DAG} allowed CGP to use smaller genomes.
In light of our current results, this conclusion should be amended.
It appears that \emph{Single} is most effective when the genome size is small and
that \emph{Reorder} and \emph{DAG} are best able to cope with smaller genome sizes.
Independent of duplication method, the ordering methods do not seem to create any
pressure on which genome size is chosen.

Unlike genome size, all configurations choose approximately the same mutation rate
for each problem.  With the exception of Parity, there appears
to be a general consensus that each gene should be mutated with probability 0.002,
independent of genome size.  Parity used a higher mutation rate, but this is likely
because 3-bit Parity is by far the easiest problem tested and therefore likely benefits
from higher exploration.

On many of the problems, many of the configurations have little to no statistical
difference from the control in their evaluations required to reach the global optimum.
On Parity and Encode it is probably safe to conclude that \emph{Single} is worse than
the control.  As these are the easiest problems tested, that difference may have
little real world meaning.  \emph{DAG} is almost certainly worse than the control on
Decode.  Its results are so much worse that we can also likely conclude that \emph{DAG}
is worse than all non \emph{DAG} configuration on that problem, even though no statistical
tests are presented here.  The control on multiply did by far the worst of all configurations,
with only \emph{Accumulate Normal} likely failing statistical significance.

By ordering the algorithm combinations by median evaluations to success (MES) and examining that ordering across each
problem, we can look for general trends.  From the table, it appears that on easier problems
the method of duplication used is the strongest predictor of rank, while
the ordering methods are more predictive on harder problems.  This comes from the
observation that in Parity and Encode (Upper half of Table~\ref{tab:results})
the first column is effectively sorted, while for Decode and Multiply
(Lower half of Table~\ref{tab:results})
the second column is sorted.
Each table includes coincidentally one exception to these rules.  On Parity and
Encode, pairings with \emph{DAG} are ranked lower than other ordering methods, so much
so as to break the ordering once on both problems.  On Decode \emph{Accumulate-Reorder}
ranks worse than expected.  Finally on Multiply \emph{Single-DAG} does far better
than expected.

@@

So now the question is why does \emph{Normal} use so little of its genome?  To
answer this we must consider what can prevent a node from becoming active.  For
a node to never be active, all tested offspring with mutations that result in that node being activated
must have been less fit than their parent.  Even if those offspring are no more fit, the selection method
used by CGP would have allowed that offspring to replace its parent, at least probabilistically.  This leads
to two possible causes for a node never being active: bad genes or infrequent node sampling.

The bad genes cause suggests the behavior encoded
by a node is simply useless on the problem landscape.  If this was the root cause
of never active nodes, we would expect \emph{Reorder} and \emph{DAG} to have
similar problems, as neither technique explicitly changes the likelihood of node
behaviors.  Furthermore, we would expect the location of never active nodes to be
uniform in the genome, which is obviously not the case.  Finally, even if bad
genes were the problem, we would expect the mutational drift allowed by CGP's
selection method to change their bad genes and eventually allow them to be active.
